Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-06-06 R/Analysis_v1.2_mash_using_greedy_covar.qmd
revision2: Add SI Appendix Sections on BLUPs, GWAS, SNP subset selection for mash

GWAS were run on BLUPs from models run at each garden individually. The GWAS were run in R/Plots_v0.3_manuscript_figures.Rmd, starting at line 1243. I copy-pasted that info here to have the full workflow in one qmd document; you may need to refer to it if you need additional information.

@@

```{r}
library(bigsnpr)
library(tidyverse)
library(switchgrassGWAS)
library(AnnotationDbi)
txdb <- loadDb(file = file.path("~", "Github", "pvdiv-genome",
                                "Pvirgatum_516_v5.1.gene.txdb.sqlite"))
workingdir <- file.path("~", "Github", "pvdiv-phenology-gxe")
inputdir <- file.path(workingdir, "analysis", "gwas", "BLUPs")
outputdir <- file.path(workingdir, "analysis", "gwas", "BLUPs")
subpop_v <- list(Gulf_and_Midwest = c("Gulf", "Midwest"), Midwest = "Midwest", Gulf = "Gulf")
snp_v <- c("12\\.1M", "8\\.8M", "10\\.3M")

@@

```{r}
#| eval: false
#| 
#|

@@

# ----------------

@@

# Step 5: Redo writeup & figures

@@

pa_legend <- plot_covar_Ulist_fig2(Ulist_gg$`Daylength (14d prior)`, legend = TRUE) + 
  labs(subtitle = "Daylength (14d prior)")
legend <- get_legend(pa_legend) 
```

>>>>

GWAS were run on BLUPs from models run at each garden individually. The GWAS were run in R/Plots_v0.3_manuscript_figures.Rmd, starting at line 1243. I copy-pasted that info here to have the full workflow in one qmd document; you may need to refer to the Rmd file if you need additional information.

### BLUPs

Copied this code from R/Plots_v0.2_manuscript_figures.Rmd, which is where the ASREML code to calculate BLUPs individually for each location x subpop x phenotype came from; you may need to refer to the Rmd file if you need additional information.

ASReml h2 greenup analysis
```{r}
#| eval: false
#| 
library(tidyverse)
library(asreml)
library(switchgrassGWAS)
library(cowplot)

source("~/Github/Functions_ggplot-theme-adjustments_2018-01-03.R")
workingdir <- file.path("~", "Github", "pvdiv-phenology-gxe")
outputdir <- file.path(workingdir, "analysis", "heritability")
metadata <- read_rds(file.path(workingdir, "data", "metadata.rds"))
sites <- read_rds(file.path(workingdir, "data", "sites.rds"))
subpops <- read_rds(file.path(workingdir, "data", "subpop_color_coding.rds"))
phe_gr_df <- read_rds(file.path(workingdir, "data",
                                "Weather_related_phe_for_asreml_h2_GR50.rds"))
phe_fl_df <- read_rds(file.path(workingdir, "data",
                                "Weather_related_phe_for_asreml_h2_FL50.rds"))
k_full <- read_rds("~/Github/pvdiv-genome/tensite_twoyear/Kinship_van_Raden_630_individuals_SNPs_r2_20percent.rds")

get BLUPs from ASReml
```{r}
site_v <- list(eight_sites = c("TX1", "TX2", "TX3", "OK", "MO", "NE", "MI", "SD"))
subpop_v <- list(Gulf = "Gulf", Midwest = "Midwest", Gulf_and_Midwest = c("Gulf", "Midwest"))
phe_fl <- c("FL50", "dyln_fl50", "dyln_change_sec", "cgdd_12c_gr2fl")
phe_gr <- c("GR50", "tmin_10d", "cgdd_12c_10d")
phe_dfs <- list(phe_gr_df, phe_fl_df)
phe_lists <- list(phe_gr, phe_fl)
h2_table <- tibble()

i = 1
j = 1
k = 1
p = 2

for(p in seq_along(phe_dfs)){
  phe_v <- phe_lists[[p]]
  phe <- phe_dfs[[p]]

  for(i in seq_along(site_v)){
    for(j in seq_along(subpop_v)){
      for(k in seq_along(phe_v)){
        phe_single <- phe %>%
    left_join(sites, by = "SITE") %>%
    filter(manu_site %in% site_v[[i]] & SUBPOP %in% subpop_v[[j]]) %>%
    rename(phenotype = eval(phe_v[k])) %>%
    mutate(PLANT_ID = as_factor(PLANT_ID),
           SITE = as_factor(SITE),
           PLOT_GL = as_factor(PLOT_GL)) %>%
    as.data.frame()

  if(length(site_v[[i]]) > 1){  # need to add site as a random factor if 
    # there is more than one site included in the dataset.
    tryCatch({
     asr_out <- asreml(phenotype ~ 1,
                       random = ~vm(PLANT_ID, k_full) +
                         ~idv(manu_site),
                       residual = ~idv(units),
                       data = phe_single, 
                       workspace = "3gb")

    h2_est <- vpredict(asr_out, h2 ~ V2/(V1+V2+V3)) %>% as_tibble() %>%
      mutate(site = names(site_v)[i],
             subpop = names(subpop_v)[j],
             phe = phe_v[k],
             bic = summary(asr_out)$bic[1],
             loglik = summary(asr_out)$loglik)
    blup <- summary(asr_out, coef=TRUE)$coef.random %>% as_tibble(rownames = "Effect")
    write_csv(blup, file.path(outputdir, paste0("blups_", names(site_v)[i], "_",
                                                 names(subpop_v)[j],
                                                 "_", phe_v[k], ".csv")))
    }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  } else {
    tryCatch({
    asr_out <- asreml(phenotype ~ 1,
                       random = ~vm(PLANT_ID, k_full),
                       residual = ~idv(units),
                       data = phe_single, 
                       workspace = "3gb")

    h2_est <- vpredict(asr_out, h2 ~ V1/(V1+V2)) %>% as_tibble() %>%
      mutate(site = names(site_v)[i],
             subpop = names(subpop_v)[j],
             phe = phe_v[k],
             bic = summary(asr_out)$bic[1],
             loglik = summary(asr_out)$loglik)
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

  if(i == 1 & j == 1 & k == 1){
    h2_table <- h2_est
  } else {
    h2_table <- add_row(h2_table, h2_est)
  }
  tryCatch({
      save_plot(filename=file.path(outputdir, paste0("ASReml_performance_",
                                                 names(site_v)[i], "_",
                                                 names(subpop_v)[j],
                                                 "_", phe_v[k], ".png")), 
            plot = plot(asr_out))
    }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})

      }
    }
  }

set up phenotype df for gwas
```{r}
#| eval: false 
#| 
site_v <- list(eight_sites = c("TX1", "TX2", "TX3", "OK", "MO", "NE", "MI", "SD"))
subpop_v <- list(Gulf = "Gulf", Midwest = "Midwest", Gulf_and_Midwest = c("Gulf", "Midwest"))
phe_fl <- c("FL50", "dyln_fl50", "dyln_change_sec", "cgdd_12c_gr2fl")
phe_gr <- c("GR50", "tmin_10d", "cgdd_12c_10d")
phe_dfs <- list(phe_gr_df, phe_fl_df)
phe_lists <- list(phe_gr, phe_fl)

i = 1
j = 1
k = 1
p = 2
phe_gwas <- tibble()

for(j in seq_along(subpop_v)){
  for(p in seq_along(phe_dfs)){
    phe_v <- phe_lists[[p]]
    phe <- phe_dfs[[p]]
    for(i in seq_along(site_v)){
      for(k in seq_along(phe_v)){
        phe_raw <- read_csv(file.path(outputdir, paste0("blups_",
                                                        names(site_v)[i], "_",
                                                        names(subpop_v)[j],
                                                        "_", phe_v[k],
                                                        ".csv")))
        names(phe_raw)[2] <- paste0(phe_v[k], "_G_BLUP")
        phe_proc <- phe_raw %>%
          separate(Effect, into = c("factor", "PLANT_ID"), sep = "\\)_") %>%
          filter(!is.na(PLANT_ID)) %>%
          left_join(metadata, by = "PLANT_ID") %>%
          filter(SUBPOP %in% subpop_v[[j]]) %>%
          select(PLANT_ID, 3)
        if(i == 1 & k == 1 & p == 1){
          phe_gwas <- phe_proc
        } else {
          phe_gwas <- phe_gwas %>%
            left_join(phe_proc, by = "PLANT_ID")
        }
      }
    }
  }
write_csv(phe_gwas, path = file.path(outputdir, paste0("G_BLUP_phenotypes_",
                                                       "for_gwas_GR_FL_",
                                                       names(subpop_v)[j],
                                                       ".csv")))
}
```

@@

NB: only the single_site BLUPs & GWAS that I generated were used in the subsequent analysis.
```{r}
#| eval: false
#| 
library(bigsnpr)
library(tidyverse)
library(switchgrassGWAS)
library(AnnotationDbi)
library(here)
txdb <- loadDb(file = here("data-raw", "ignore",
                           "Pvirgatum_516_v5.1.gene.txdb.sqlite"))
workingdir <- file.path("~", "Github", "pvdiv-phenology-gxe")
inputdir <- file.path(workingdir, "analysis", "gwas", "BLUPs")
outputdir <- file.path(workingdir, "analysis", "gwas", "BLUPs")
subpop_v <- list(Gulf_and_Midwest = c("Gulf", "Midwest"), Midwest = "Midwest", Gulf = "Gulf")
snp_v <- c("12\\.1M", "8\\.8M", "10\\.3M")

@@

```{r}
#| eval: false
#|

@@

@@

# ---
# Step 5: Redo writeup & figures

@@

pa_legend <- plot_covar_Ulist_fig2(Ulist_gg$`Daylength (14d prior)`, legend = TRUE) + 
  labs(subtitle = "Daylength (14d prior)")
legend <- get_legend(pa_legend)

### 2024-06-06 R/SI_Appendix_Datasets.Rmd
revision2: Add SI Appendix Sections on BLUPs, GWAS, SNP subset selection for mash

    Dataset 1. SNP-associated effects and standard errors for green-up estimated using the Gulf genetic subpopulation.
    Dataset 2. SNP-associated effects and standard errors for green-up estimated using the Midwest genetic subpopulation.
    Dataset 3. SNP-associated effects and standard errors for green-up estimated using Both genetic subpopulations.
    Dataset 4. SNP-associated effects and standard errors for flowering estimated using the Gulf genetic subpopulation.
    Dataset 5. SNP-associated effects and standard errors for flowering estimated using the Midwest genetic subpopulation.
    Dataset 6. SNP-associated effects and standard errors for flowering estimated using Both genetic subpopulations.
    Dataset 7. Genes within QTL that have functionally validated roles in flowering in rice.
    Dataset 8. Overlap between quantitative trait loci for flowering measured using an outbred pseudo-F2 mapping population, and significant mash effect estimates (with a log10Bayes Factor > 2, or in the 1% tail of most significant mash results) found using three overlapping diversity panels. Genes within 20kb with a functionally validated role in flowering in rice are  indicated.

@@

@@

Find this file on NAS: "~/Github/pvdiv-genome/Pvirgatum_516_v5.1.gene.txdb.sqlite"

@@

# Dataset 8
All flowering and green-up QTL intervals contained at least one SNP significant in at least one mash run at a log10-transformed Bayes Factor > 2, or in the 1% tail of significance, whichever was stricter.

## Objects generated earlier
Get `qtlplot` object by running code starting at about line 150 (for Dataset 7).
Get `m_gulf_fl_hyp` and other objects by running code starting at about line 35 (for Datasets 1-6).

## Thresholds
Decide on cutoff to use for rug plots and enrichment analysis: SNPs sig in 1% tail or BF\>2, whichever is stricter

>>>>

```         
Dataset 1. SNP-associated effects and standard errors for green-up estimated using the Gulf genetic subpopulation.
Dataset 2. SNP-associated effects and standard errors for green-up estimated using the Midwest genetic subpopulation.
Dataset 3. SNP-associated effects and standard errors for green-up estimated using Both genetic subpopulations.
Dataset 4. SNP-associated effects and standard errors for flowering estimated using the Gulf genetic subpopulation.
Dataset 5. SNP-associated effects and standard errors for flowering estimated using the Midwest genetic subpopulation.
Dataset 6. SNP-associated effects and standard errors for flowering estimated using Both genetic subpopulations.
Dataset 7. Genes within QTL that have functionally validated roles in flowering in rice.
Dataset 8. Overlap between quantitative trait loci for flowering measured using an outbred pseudo-F2 mapping population, and significant mash effect estimates (with a log10Bayes Factor > 2, or in the 1% tail of most significant mash results) found using three overlapping diversity panels. Genes within 20kb with a functionally validated role in flowering in rice are  indicated. 
```

@@

## SI Appendix Table S1 Info

#### random SNP inputs

```{r}
popkey = list(all = "Gulf_and_Midwest", midwest = "Midwest", gulf = "Gulf")
phekey = list(greenup = "GR50", flowering = "FL50")
# i=2; j=1

for (i in seq_along(popkey)) {
  for (j in seq_along(phekey)) {

    if (i == 2 & j == 1) {
      betas <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("B_hat_random_df_24000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
      stderr <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("S_hat_random_df_24000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
    } else if (i == 2 & j == 2) {
      betas <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("B_hat_random_df_33000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
      stderr <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("S_hat_random_df_33000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
    } else {
      betas <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("B_hat_random_df_19000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
      stderr <- readRDS(file = here("analysis", "mash", popkey[[i]], "inputs", 
                                 paste0("S_hat_random_df_19000topSNPs_",
                                        phekey[[j]], "_", popkey[[i]], ".rds")))
    }
    print(paste(names(popkey)[i], names(phekey)[j], "random SNP set (input into greedy mash algorithm) has", nrow(betas), "SNPs"))
  }
}

```

So random SNP sets have 71-127K SNPs.

#### strong SNP inputs

Numbers for this section: We then selected 19K relatively unlinked (r2 < 0.2) SNPs from each univariate GWAS that had the largest effect estimates in each GWAS (hereafter "strong" effects), which gave sets of 48K - 80K SNPs, and jointly re-estimated these effect size estimates at all eight sites using the set of significant covariance matrices that most improved the model fit on the random effects. Notably, the majority of these "strong" effects were not significant in univariate GWAS.

```{r}
switchgrassGWAS:::get_marker_df(m_gulf_gr_hyp) # 45931
switchgrassGWAS:::get_marker_df(m_gulf_fl_hyp) # 50490
switchgrassGWAS:::get_marker_df(m_midw_gr_hyp) # 49030
switchgrassGWAS:::get_marker_df(m_midw_fl_hyp) # 63367
switchgrassGWAS:::get_marker_df(m_gam_gr_hyp)  # 44765
switchgrassGWAS:::get_marker_df(m_gam_fl_hyp)  # 35735
```

@@

Find this file on NAS: "\~/Github/pvdiv-genome/Pvirgatum_516_v5.1.gene.txdb.sqlite"

@@

# Dataset 8

All flowering and green-up QTL intervals contained at least one SNP significant in at least one mash run at a log10-transformed Bayes Factor \> 2, or in the 1% tail of significance, whichever was stricter.

## Objects generated earlier

Get `qtlplot` object by running code starting at about line 150 (for Dataset 7). Get `m_gulf_fl_hyp` and other objects by running code starting at about line 35 (for Datasets 1-6).

## Thresholds

Decide on cutoff to use for rug plots and enrichment analysis: SNPs sig in 1% tail or BF\>2, whichever is stricter

### 2024-06-06 manuscript/revision2/SI_Appendix/SI_Appendix.qmd
revision2: Add SI Appendix Sections on BLUPs, GWAS, SNP subset selection for mash

References cited in the SI text should be included in a separate reference list at the end of this SI file: @varga2016multilingual and @olsen1992optimal.

Supporting information for Brief Reports is limited to extended methods, essential supporting datasets, and videos (no additional tables or figures). Supporting figures and tables are not allowed for Brief Reports.

@@

## Subhead {.unnumbered}

Type or paste text here. This should be additional explanatory text such as an extended technical description of results, full details of mathematical models, etc.

@@

## Section S1. Covariance matrices generated to jointly re-estimate SNP effects across eight sites {.unnumbered}

@@

We were interested in the types of genetic correlation that could be expressed by different alleles across the genome and reasoned that different loci could have different patterns of genetic correlation. For example, one locus's effects could covary with a weather-based cues, if it had an allele that was responsive to this cue. Alternatively, a locus might have an effect at only one garden, for example if a herbivore or pathogen was only present at one garden.

We specified three qualitative categories of covariance structures across gardens: "canonical" covariance, with simple patterns of effect size covariance introduced in the initial *mash* manuscript; "data-driven" covariances derived from common patterns of SNP effects observed in the data and introduced in the initial *mash* manuscript, and "GxWeather" covariance, estimated from the covariance of empirical weather patterns at each garden over specific time frames before the phenological event.

@@

Data-driven covariance matrices can also be created by *mash* and fall into two groups: an overall pattern of covariance between all conditions, denoted as $tPCA$, and the first five principle components of a singular value decomposition of the $tPCA$, denoted as `PCA_1`$PCA_1$ through $PCA_5$[@Urbut2019].

To create the GxWeather covariance matrices we transformed each genotype's green-up and flowering date at each garden to create date-related values that were summary functions for a weather variable across a specified date range. For example, the cumulative sum of the rainfall that occurred seven days prior to the flowering date was used to create one possible GxWeather cue variable for flowering date. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 208.

We defined both phenological dates as functions of five weather variables and eight to ten time frames (Table 1): Cumulative growing degree days (GDD) variables for each time frame were calculated as $GDD = \sum^{DATE}_{DATE - i} max(T_{mean} - T_{base}, 0)$, where $T_{mean}$ is the daily average temperature, defined as $(T_{max} + T_{min}) / 2$, $T_{base}$ is the base temperature of 12°C for switchgrass, $T_{max}$ is the maximum daily temperature, $T_{min}$ is the minimum daily temperature, $DATE$ is the phenological date, and $i$ is the time frame, or the number of days prior to the phenological date that GDD is summed across [@kiniry2005; @behrman2013]. Second, cumulative rainfall values in each time frame were calculated as $C_{rain} = \sum^{DATE}_{DATE - i} PRCP$, where PRCP is the daily rainfall measured in millimeters, $DATE$ is the phenological date, and $i$ is the time frame, or the number of days prior to the phenological date that rainfall is summed across. Third, day length in hours was determined for each time frame as a specific single day prior to the phenological date, as indicated by the time frame $i$, e.g. $DATE - i$. Day length was calculated as a function of latitude and day of the year as in [@forsythe1995]. Fourth, day length change in seconds was determined for a specific single day prior to the phenological date as indicated by the time frame, as $DYLN_{change (s)} = (DYLN_{DATE - i} - DYLN_{DATE-i-1})*60*60$, where $DYLN$ in the day length in hours, $DATE$ is the phenological date, and $i$ is the number of days prior to the phenological date. Fifth, average temperature in each time frame was defined as $T_{ave} = \sum^{DATE}_{DATE - i} (T_{max} + T_{min}) / 2$, where $T_{max}$ is the maximum daily temperature, $T_{min}$ is the minimum daily temperature, $DATE$ is the phenological date, and $i$ is the time frame, or the number of days prior to the phenological date that average temperature is summed across.

We then generated GxWeather covariance matrices derived from correlations in these GxWeather cue variables for the three subpopulations. These covariance matrices represent the correlations between genotypes for these weather-related variables across our eight common gardens. To create these matrices, we determined the correlations between these weather-related cue variables for identical genotypes grown at different common gardens and used these correlations to fill the off-diagonals of a eight-by-eight common garden matrix. For the diagonal of these matrices, we used the narrow-sense heritability for each weather-related cue variable at each garden, calculated using the same methods as Section S3. Matrices need to be positive semi-definite to be used in *mash*, so when the resultant matrices were not positive semi-definite, the `Matrix::nearPD` function was used to compute the nearest positive definite matrix to the correlation matrix [@Matrix]. Finally, all covariance matrices are rescaled so that the maximum diagonal element of the rescaled matrix is 1, as in [@Urbut2019]. Code used to generate these matrices can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 536.

@@

## Section S2. Greedy mash algorithm used to select covariance matrices that significantly improved the model log-likelihood {.unnumbered}

We generated 48 - 60 GxWeather covariance matrices for each of our five weather variables over different time frames. These GxWeather covariance matrices were in many cases highly correlated.

selection of the GxWeather

> describing that the weather-based cues are themselves correlated, and picking and choosing among weather cues is a model building problem – to address the issue, we used a heuristic greedy algorithm to pick the weather-based covariance structures that improved the likelihood of the mash model

> A lot of the results interpretation is based on interpreting the SNP loadings on the specified covariance matrices. This is a secondary use of mash (the primary being the refinement of effect sizes), and while they do interpret these somewhat in the mash paper, these loadings need to be interpreted with caution. If these covariance matrices are similar, mash will somewhat arbitrarily assign weight to any of the matrices because all lead to the same fit to the effect-size data. The mash algorithm doesn’t give the full posterior distribution on the loadings so you can’t check for posterior correlations there. It seems likely to me that the hypothesis-driven and data-driven covariance matrices are somewhat correlated here, and the correlations may differ between green-up and flowering because of the better correlation between the environmental metrics and flowering than green-up.

Type or paste text here. Break this section up into subheads as needed

## Section S3. Narrow-sense heritability estimation {.unnumbered}

@@

## Section S4. Outbred pseudo-F2 mapping population and Quantitative Trait Locus mapping {.unnumbered}

@@

| Species           |  CBS |   CV |   G3 |
|:------------------|-----:|-----:|-----:|
| Acetaldehyde      |  0.0 |  0.0 |  0.0 |
| Vinyl alcohol     |  9.1 |  9.6 | 13.5 |
| Hydroxyethylidene | 50.8 | 51.2 | 54.0 |

: Simple table example {#tbl-simple}

@@

\dataset{Dataset_8_blah.csv}{Overlap between quantitative trait loci (QTL) for the start of vegetative or reproductive growth with significant mash effect estimates (with a log10(Bayes Factor) > 2) found using diversity panels of the Midwest, Gulf, and Both genetic subpopulations. The first five columns of this table provide QTL information, and columns 6-10 provide mash Marker information for markers within the QTL regions.}

>>>>

@@

@@

## Section S1. Covariance matrices generated to provide multiple ways SNP effects could covary across common gardens {.unnumbered}

@@

We were interested in the types of genetic correlation that could be expressed by different alleles across the genome and reasoned that different loci could have different patterns of genetic correlation. For example, the effects of a locus, $A$, could covary with a weather-based cue if $A$ had a common allele, $a$, that was responsive to this cue. In contrast, the effects of a different locus $B$ might have an effect at only one garden, for example if a pathogen was only present at one garden, and $B$ has a common allele $b$ that is resistant to that pathogen.

We specified three qualitative categories of covariance structures across gardens: "canonical" covariance, with simple patterns of effect size covariance introduced in the initial *mash* manuscript; "data-driven" covariances derived from common patterns of SNP effects observed in the data and introduced in the initial *mash* manuscript, and "GxWeather" covariance, estimated from the covariance of empirical weather patterns at each garden over specific time frames before each genotype underwent the phenological event.

@@

Data-driven covariance matrices can also be created by *mash* and fall into two groups: an overall pattern of covariance between all conditions, denoted as $tPCA$, and the first five principle components of a singular value decomposition of the $tPCA$, denoted as $PCA_1$ through $PCA_5$ [@Urbut2019].

To create the GxWeather covariance matrices we transformed each genotype's green-up and flowering date at each garden to create date-related values that were summary functions for a weather variable across a specified date range. For example, the cumulative sum of the rainfall that occurred seven days prior to the flowering date was used to create one possible GxWeather cue variable for flowering date. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 390.

We defined the two phenological dates as functions of five weather variables and eight to ten time frames (Table 1). First, cumulative growing degree days (GDD) variables for each time frame were calculated as $GDD = \sum^{DATE}_{DATE - i} max(T_{mean} - T_{base}, 0)$, where $T_{mean}$ is the daily average temperature, defined as $(T_{max} + T_{min}) / 2$, $T_{base}$ is the base temperature of 12°C for switchgrass, $T_{max}$ is the maximum daily temperature, $T_{min}$ is the minimum daily temperature, $DATE$ is the phenological date, and $i$ is the time frame, or the number of days prior to the phenological date that GDD is summed across [@kiniry2005; @behrman2013]. Second, cumulative rainfall values in each time frame were calculated as $C_{rain} = \sum^{DATE}_{DATE - i} PRCP$, where PRCP is the daily rainfall measured in millimeters, $DATE$ is the phenological date, and $i$ is the time frame, or the number of days prior to the phenological date that rainfall is summed across. Third, day length in hours was determined for each time frame as a specific single day prior to the phenological date, as indicated by the time frame $i$, e.g. $DATE - i$. Day length was calculated as a function of latitude and day of the year as in [@forsythe1995]. Fourth, day length change in seconds was determined for a specific single day prior to the phenological date as indicated by the time frame, as $DYLN_{change (s)} = (DYLN_{DATE - i} - DYLN_{DATE-i-1})*60*60$, where $DYLN$ in the day length in hours, $DATE$ is the phenological date, and $i$ is the number of days prior to the phenological date. Fifth, average temperature in each time frame was defined as $T_{ave} = \sum^{DATE}_{DATE - i} (T_{max} + T_{min}) / 2$, where $T_{max}$ is the maximum daily temperature, $T_{min}$ is the minimum daily temperature, $DATE$ is the phenological date, and $i$ is the time frame, or the number of days prior to the phenological date that average temperature is summed across.

We then generated GxWeather covariance matrices derived from correlations in these GxWeather cue variables for the three subpopulations. These covariance matrices represent the correlations between genotypes for these weather-related variables across our eight common gardens. To create these matrices, we determined the correlations between these weather-related cue variables for identical genotypes grown at different common gardens and used these correlations to fill the off-diagonals of a eight-by-eight matrix. To fill the diagonal entries of these matrices, we used the narrow-sense heritability for each weather-related cue variable at each garden, calculated using the same methods as Section S3. Matrices need to be positive semi-definite to be used in *mash*, so when the resultant matrices were not positive semi-definite, the `Matrix::nearPD` function was used to compute the nearest positive definite matrix to the correlation matrix [@Matrix]. Finally, all covariance matrices were rescaled so that the maximum diagonal element of the rescaled matrix is 1, as in [@Urbut2019]. Code used to generate these matrices can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 716.

@@

## Section S2. Univariate genome-wide association to obtain initial effect estimates {.unnumbered}

To obtain initial effect estimates for each single nucleotide polymorphism (SNP) for each combination of phenological trait, genetic subpopulation, and common garden, we conducted univarate genome-wide association (GWAS) on site-specific best linear unbiased predictors (BLUPs) using the switchgrassGWAS package and the methods in [@Lovell2021]. The full SNP diversity panel generated in [@Lovell2021] was subset within each subpopulation for all subsequenct analyses to retain only SNPs with $\leq$ 20% missing data and minor allele frequencies \> 0.05. This subsetting resulted in 8.8 million SNPs retained for the Midwest subpopulation, 10.3 million SNPs retained for the Gulf subpopulation, and 12.3 million SNPs retained for Both subpopulations.

BLUPs were calculated independently for each combination of trait, genetic subpopulation, and common garden. We used ASReml to estimate BLUPs using mixed models of the form:

$$
Var(u) = G\sigma^2_u
$$

$$
Var(e) = I\sigma^2_e
$$

where the vector $\mathbf{y}$ represents the flowering date or greenup date values for that garden, $Z$ the design matrix for random effects, $u$ the whole genome additive genetic effect, $e$ the residual, $G$ the genomic relationship matrix based on all SNPs retained for subpopulation-specific analyses, and $I$ the rank-$\mathbf{y}$ identity matrix [@clark2013]. ASReml can specify considerably more complex variance structures; however, we were interested in specifying covariance structures using *mash*, which allows the specification of multiple covariance structures; hence, we used a relatively simple model to calculate BLUPs for use in univariate GWAS. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 74.

Univariate GWAS were run independently for each combination of trait, genetic subpopulation, and common garden. The code used to run GWAS can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 240. We used [`pvdiv_standard_run()`](https://alice-macqueen.github.io/switchgrassGWAS/reference/pvdiv_standard_gwas.html), an opinionated function of the [switchgrassGWAS](https://github.com/Alice-MacQueen/switchgrassGWAS) R package, to conduct GWAS. To control for population structure and run GWAS, this function uses a three-part process. First, a fast truncated singular value decomposition (SVD) is used, which has initial pruning of regions in long-range linkage disequilibrium (LD) and iterative pruining and removal of regions in long-range LD [@privé2017a]. By default, 15 vectors of singular values, $k$, are computed. Pruning and removal of long-range LD regions are recommended in human genetics to best control for population structure [@abdellaoui2013; @privé2017; @price2008]. Second, linear regression models using 0-15 of these vectors of singular values were used to run GWAS and to calculate the genomic control coefficient, $\lambda_{GC}$, for each $k$ [@Devlin1999]. Third, we chose the lowest value of $k$ that made $\lambda_{GC}$ closest to 1 to correct for population structure, and ran GWAS using a linear regression model with this $k$ for population structure correction.

## Section S3. Selection of effect estimate subsets to model using *mash* {.unnumbered}

To make the subsequent analyses using *mash* computationally feasible, we followed the author's [recommended analysis outline](https://stephenslab.github.io/mashr/articles/eQTL_outline.html) for large datasets using *mash* and considered subsets of the millions of effects at different stages of the mash analysis. We extracted two subsets of tests from our complete GWAS effect estimate datasets: 1) effects from a subset of "strong" tests corresponding to stronger effects on our traits; 2) results from a random subset of all tests to correspond to an unbiased representation of all effects. We also clumped effects before selection so that signal strength was standardized across different LD blocks in the genome, and not limited to closely linked associations.

To prepare subsets of effect estimates and their standard errors from these GWAS for use in mash, we used [`pvdiv_bigsnp2mashr()`](https://alice-macqueen.github.io/switchgrassGWAS/reference/pvdiv_bigsnp2mashr.html) from the switchgrassGWAS R package. For each phenological trait and each genetic subpopulation, we created two subsets of effects and standard errors in paired $m \times n$ matrices, where $m$ were marker effects and $n$ were common gardens. The first subset of effects and standard errors, the strong effects, was created by selecting relatively unlinked (r^2^ \< 0.2) SNPs from each univariate GWAS that had the smallest *p*-values in those GWAS. Notably, the majority of these "strong" effects were not significant at a 5% level in univariate GWAS. The number of SNPs from each univariate GWAS was chosen to result in approximately million entry matrices for each trait & subpopulation. We selected 19K - 33K effects and standard errors for the strong effect subsets (@tbl-simple). In many cases, there was overlap in the markers with the largest effect estimates at different locations; thus, 19 - 33K SNPs selected for each of eight univariate GWAS ultimately gave sets of 35K - 63K SNPs (@tbl-simple). The second subset of effects and standard errors, the random subset, was created by selecting relatively unlinked (r^2^ \< 0.2) SNPs from each univariate GWAS after clumping using minor allele frequency. To account for the chance that true effects could be sparse, we doubled the size of the random subset relative to the strong subset. We selected 38 - 66K SNP effects and formed subsets of 71 - 127K SNP markers for the random subset.

## Section S4. Greedy *mash* algorithm used to select covariance matrices that significantly improved the model log-likelihood {.unnumbered}

Because we were particularly interested in determining whether genomic loci varied in their patterns of genetic covariance across gardens, specifying the optimal set of covariance matrices in *mash* was an important model-building problem. If very similar covariance matrices are included when fitting a model in *mash*, *mash* will arbitrarily assign effect loadings across these matrices, as all assignments will lead to the same fit to the effect-size data. Our GxWeather covariance matrices were in many cases highly correlated, as they represented covariation in the same weather variables in overlapping time frames. We thus evaluated the performance of *mash* models when different covariance matrices were included by implementing a "greedy" mash algorithm.

## Section S5. *Mash* to jointly re-estimate SNP effects across eight common gardens {.unnumbered}

## Section S6. Narrow-sense heritability estimation {.unnumbered}

@@

## Section S7. Outbred pseudo-F2 mapping population and Quantitative Trait Locus mapping {.unnumbered}

@@

| Phenology type | Subpopulation |  Effects Selected per GWAS, Strong Subset |  Effects Selected per GWAS, Random Subset | Total Markers, Strong Subset | Total Markers, Random Subset |
|:-----------|--------------:|-----------:|-----------:|------------|------------|
| Greenup Date   |          Gulf |                                     19000 |                                     38000 | 45931                        | 91862                        |
|                |       Midwest |                                     24000 |                                     48000 | 49030                        | 98060                        |
|                |          Both |                                     19000 |                                     38000 | 44765                        | 89530                        |
| Flowering Date |          Gulf |                                     19000 |                                     38000 | 50490                        | 100980                       |
|                |       Midwest |                                     33000 |                                     66000 | 63367                        | 126734                       |
|                |          Both |                                     19000 |                                     38000 | 35735                        | 71470                        |

: Effect estimates selected for joint re-estimation across eight common gardens using *mash* {#tbl-simple}

@@

\dataset{Dataset_8_QTL_mash_overlap.csv}{Overlap between quantitative trait loci (QTL) for the start of vegetative or reproductive growth with significant mash effect estimates (with a log10(Bayes Factor) > 2) found using diversity panels of the Midwest, Gulf, and Both genetic subpopulations. The first five columns of this table provide QTL information, and columns 6-10 provide mash Marker information for markers within the QTL regions.}
